---
title: "HW2"
author: "Solutions"
output: pdf_document
---

The code below just loads some packages and makes it so that enough digits are printed that you won't get confused by rounding errors.

```{r, message = FALSE, warning=FALSE}
library(dplyr) # functions like summarize
library(ggplot2) # for making plots
library(mosaic) # convenient interface to t.test function

options("pillar.sigfig" = 10) # print 10 significant digits in summarize output
```

## Problem 1: Adapted from Sleuth3 2.20

Researchers assigned 7 volunteer males to a special fish oil diet.  They recorded each subject's diastolic blood pressure at baseline and again after four weeks on the diet.  The researcher's interest was in the reduction diastolic blood pressure (mm of mercury) between baseline and 4 weeks later (a positive reduction is a good thing).  The R code below reads in the data set.  The variable `BP` records each subject's reduction in blood pressure.


```{r}
fish_oil <- read.csv("http://www.evanlray.com/data/sleuth3/ex0112_oil_diets.csv") %>%
  filter(Diet == "FishOil")
```

### (a) Define the relevant parameter(s) the researchers wanted to estimate.

The parameter of interest is the mean reduction in diastolic blood pressure after 4 weeks on a fish oil diet ($\mu$), in a population of people similar to the seven volunteers in the study.

Be sure to address the scope of your conclusions when you draw conclusions for the hypothesis test and interpret the confidence interval.

### (b) Calculate the sample mean and standard deviation of the blood pressure reduction.

```{r}
fo_summary <- 
fish_oil %>%
  summarise(
    mean_bp = mean(BP),
    sd_bp = sd(BP)
  )

fo_summary
```


### (c) Compute the standard error for the sample mean.  What are the degrees of freedom?

```{r}
fo_se <- fo_summary$sd_bp/sqrt(7)
fo_se
```

The degrees of freedom is 7-1=6.

### (d) Conduct a relevant hypothesis test.

#### i. Define the null and alternative hypotheses for the test

$H_0: \mu=0$. On average, the fish oil diet has no impact on a person's blood pressure, in the population of people who are like those who were volunteers in this study.

$H_A: \mu\neq 0$. On average, there is some impact of the fish oil diet on a person's blood pressure, in the population of peole who are like the volunteers in this study.

Note, if you went into the study (before collecting data) believing that fish oil would reduce blood pressure, then your alternative might be $H_A: \mu >0$. In general in this class we have just been using two-sided alternatives, but this is a situation where you might you a one-sided hypothesis ONLY if you set the hypotheses before collecting data.

#### ii. Find the value of the $t$ statistic for this test.   What is the interpretation of this statistic?

```{r}
tstat <- fo_summary$mean_bp/fo_se
tstat
```

The sample mean reduction in blood pressure after 4 weeks on the fish oil diet is almost 3 standard errors larger than the difference of 0 specified in the null hypothesis.

#### iii. Find the p-value for the test using the `t.test` function in R.  Don't forget to specify the value of `mu` from the null hypothesis.  (See the corrected handout on the course website from Mon, Sep 9 -- the original handout had incorrect code.)

```{r}
t.test(~BP, mu=0, data=fish_oil)
```

The p-value for a two-sided hypothesis is 0.02498. 

#### iv. What is the interpretation of the p-value?  (I'm not looking for a conclusion about strength of evidence yet, just a statement of what the p-value is in the context of this example.)

If you used a two-sided hypothesis test:

If on average the fish oil diet had no impact on a person’s blood pressure, the probability of observing a t statistic at least as large (in abolute value) as 2.9693 is 0.02498.

#### v. What is the conclusion of the test?  Please state this in terms of strength of evidence about the null hypothesis.

The data provide a moderate amount of evidence against the null hypothesis that fish oil has no effect on a person’s blood pressure, in the population of people similar to those who volunteered to participate in the study.

### (e) Find a relevant confidence interval.

#### i. Find a confidence interval from the formula, using output from the `qt` function as needed.  Confirm that your interval matches the interval from the output of `t.test` in part (d) iii, up to rounding error.

```{r}
tcrit <- qt(0.975, df=6)

lb <- fo_summary$mean_bp - tcrit*fo_se
ub <- fo_summary$mean_bp + tcrit*fo_se

c(lb, ub)
```

The confidence interval calculated here is (1.156, 11.987), which matches the interval we got from t.test.

#### ii. Interpret your interval in context, including a statement of what the phrase "95% confident" means.

We are 95% confident that the mean change in a person's blood pressure after 4 weeks on a fish oil diet, among the population of people similar to those who participated in this study, is between 1.156 and 11.987. If we were to take many different samples from this population and calculate a similar 95% confidence interval based on each of those samples, approximately 95% of those confidence intervals would contain the mean change in a person's blood pressure after 4 weeks on a fish oil diet, among the population of people similar to those who participated in this study.

#### iii. Is the confidence interval from part ii guaranteed to contain the sample mean?

Yes: the interval is obtained by subtracting a margin of error from the sample mean and adding a margin of error to the sample mean, so the sample mean is guaranteed to be contained in the confidence interval.

#### iv. Is the confidence interval from part ii guaranteed to contain the population mean?

No. For 95% of samples, a confidence interval calculated based on that sample will contain the population mean. However, for 5% of samples, a confidence interval calculated based on that sample will not contain the population mean. For a given sample, there is no way to know whether or not the confidence interval we have calculated contains the population parameter.



## Problem 2: Adapted from Sleuth3 2.23

Ther National Highway System Designation Act was signed into law in the United States on November 29, 1995.  Among other things, the act abolished the federal mandate of 55-mile-per-hour maximum speed limits on roads in the United States and permitted states to establish their own limits.  Of the 50 states (plus the District of Columbia), 32 increased their speed limits either at the beginning of 1996 or sometime during 1996.

The R code below reads in data with the percentage changes in interstate highway traffic fatalities from 1995 to 1996 (the variable is called `PctChange` in the data frame).  Among the states where the speed limit increased, what evidence is there that the average percent change in fatalities was different from 0?

Conduct a full analysis, including:

 * an appropriate plot with informative axis labels,
 * summary statistics that would be meaningful to someone who had not taken a statistics class (i.e., don't report the t statistic),
 * a confidence interval, and
 * a hypothesis test.

Interpret all of your results in context.  Explain how to interpret the p-value for the test and the conclusions that can be drawn from it as though to someone who had not taken a statistics class.  Similarly, explain how to interpret your confidence interval.  You do not need to calculate the p-value or the confidence interval by hand; you can use output from the `t.test` function.  What conclusions can be drawn about whether this policy change was a good idea?

```{r}
fatalities <- read.csv("http://www.evanlray.com/data/sleuth3/ex0223_highway_safety.csv") %>%
  filter(SpeedLimit == "Inc")
```

```{r}
dim(fatalities)
```

```{r}
fatalities %>%
  summarise(
    mean_pct_change = mean(PctChange),
    min_pct_change = min(PctChange),
    max_pct_change = max(PctChange)
  )
```

```{r}
t.test(~ PctChange, data=fatalities)
```

```{r}
(ggplot(data = fatalities, aes(x = PctChange)) 
   + geom_histogram(bins = 10)
   + xlab("Percent Change in Fatalities") 
   + ylab("Number of States"))
```

This is an observational study, so it is hard to use this data set to make any definitive claims about a causal relationships between speed limits and changes in the numbers of fatalities due to traffic accidents. Additionally, we cannot really frame the states in our data set as a random sample from any population. However, these data could lend insight into the potential implications of a state increasing its speed limit under similar circumstances to the states in this sample.

Among the 32 states that increased their speed limits, there was a roughly even split between states that had a reduced frequency of fatalities due to traffic accidents and states that had increased frequency of fatalities. Across these states, the mean percent change in fatalities was about 0.5%, with a 95% confidence interval of approximately -2.4% to 3.4%. We can think of this interval as a range of plausible values for the average percent change in fatalities due to traffic accidents that we might see if a state similar to those in the sample increased its speed limit. In 95% of samples, an interval constructed in a similar way would contain the mean percent change in fatalities.

A hypothesis test of the claim that the speed limit increases had no association with changes in the number fatal accidents had a p-value of 0.73. This means that if in fact the speed limit increases were not associated with any change in the frequency of fatal accidents, there would be a 73% chance of seeing an average difference of at least 0.5% (or -0.5%). Our data are therefore consistent with that hypothesis of no effect; the data do not offer any evidence to rule out the possibility that the speed limit increase had no association with a change in the frequency of fatal accidents.

Overall, the data suggest that if the speed limit increase had any effect on the frequency of fatal accidents, it was small in percentage terms.